# Reproducible Research: Course Project 1
========================================================
#### date created: "`r Sys.Date()`"  

#### Load libraries
```{r, echo-TRUE, error=FALSE, warning=FALSE, message=FALSE, results="hide"}
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data

#### Download and unzip data file if necessary
```{r}
if(!exists("activity")) {
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  destinfile <- "./repdata_data_activity.zip"
  download.file(url, destinfile, method="curl")
  unzip(destinfile)
}
```

#### Load data and convert data variable
```{r}
activity <- read.csv("activity.csv", na.strings="NA")
activity$date <- as.Date(as.character(activity$date), format="%Y-%m-%d")
```

## What is mean total number of steps taken per day?

#### Calculate the total number of steps taken per day 
(note: missing values were omitted from calculation)
```{r}
total.steps <- activity %>% group_by(date) %>% summarize(total = sum(steps, na.rm=TRUE))
```

#### Make a histogram of the total number of steps taken each day
```{r hist_steps, warning=FALSE}
hist.steps <- ggplot(total.steps, aes(total))
hist.steps + geom_histogram(stat="bin", fill="steelblue", bins=25) + ggtitle("Frequency of the Total Number of Steps Taken Each Day") + labs(x="Total Steps", y="Count")
```

#### Calculate the mean and median total number of steps taken per day
```{r}
mn <- mean(total.steps$total)
med <- median(total.steps$total)
```
The mean and median total number of steps taken per day is `r mn` and `r med` respectively.

## What is the average daily activity pattern?

#### Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r plot_interval, warning=FALSE}
time.series <- ggplot(activity, aes(interval, steps))
time.series + stat_summary(fun.y="mean", geom="line", color="steelblue", lwd=0.75) + ggtitle("Average Number of Steps Taken Across All Days") + labs(x="5-Minute Interval", y="Mean Steps")
```

#### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
mean.ordered <- activity %>% group_by(interval) %>% summarize(mean = mean(steps, na.rm=TRUE)) %>% arrange(desc(mean))
maximum <- mean.ordered$interval[1]
```
The `r maximum` minute interval contains the maximum number of steps, averaged across all days.

## Imputing missing values

There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

#### Calculate and report the total number of missing values in the dataset
```{r}
na.sum <- sum(is.na(activity$steps))
```
There are a total number of `r na.sum` rows with NAs in the dataset.

#### Create a new dataset that is equal to the original dataset but with the missing data filled in

Assuming that this individual has a regular schedule and does similar activities at the same time everyday, all of the missing values are filled in with the mean value for that 5-minute interval.

```{r}
mean.interval <- aggregate(steps~interval, data=activity, mean)
activity.complete <- activity
activity.complete$steps <- ifelse(is.na(activity.complete$steps), mean.interval$steps[mean.interval$interval %in% activity.complete$steps], activity.complete$steps)
```


#### Make a histogram of the total number of steps taken each day
```{r hist_steps_impute}
total.steps.complete <- activity.complete %>% group_by(date) %>% summarize(total=sum(steps, na.rm=TRUE))
hist.steps.complete <- ggplot(total.steps.complete, aes(total))
hist.steps.complete + geom_histogram(stat="bin", fill="steelblue", bins=25) + ggtitle("Frequency of the Total Number of Steps Taken Each Day") + labs(x="Total Steps", y="Count")
```

#### Calculate the mean and median total number of steps taken per day
```{r}
mn.complete <- mean(total.steps.complete$total)
med.complete <- median(total.steps.complete$total)

mn.diff <- mn.complete - mn
med.diff <- med.complete - med
```
The mean and median total number of steps taken per day is `r mn.complete` and `r med.complete` respectively. After imputing data for the missing values, the mean is higher (difference of `r mn.diff`), while the median remains the same (difference of `r med.diff`).

## Are there differences in activity patterns between weekdays and weekends?

#### Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" 
(note: the imputed dataset was used)
```{r}
activity.complete$day <- ifelse(weekdays(activity.complete$date)=="Saturday"| weekdays(activity.complete$date)=="Sunday", "Weekends", "Weekdays")
```

#### Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r plot_interval_days}
wk.series <- ggplot(activity.complete, aes(interval, steps))
wk.series + stat_summary(fun.y="mean", geom="line", color="steelblue", lwd=0.75) + facet_grid(day ~.) + ggtitle("Average Number of Steps Taken Across Weekdays and Weekends") + labs(x="Interval", y="Mean Steps")
```

